Evaluating Cloud-Optimized HDF5 for NASA's ICESat-2 Mission

### 2024-12-10 _tex/paper.tex
Built site for gh-pages

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\includegraphics{figures/figure-1.png}

@@

\includegraphics{figures/figure-2.png}

@@

\includegraphics{figures/figure-3.png}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1589}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.3444}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1457}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0993}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1060}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
prefix
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\% file size increase
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textasciitilde km per chunk
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
shape
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
page size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
avg\_chunk\_size
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
original & original file from ATL03 v6 (1gb and 7gb) & 0 & 1.5km &
(10000,) & N/A & 35kb \\
original-kerchunk & kerchunk sidecar of the original file & N/A & 1.5km
& (10000,) & N/A & 35kb \\
page-only-4mb & paged-aggregated file with 4mb per page &
\textasciitilde1\% & 1.5km & (10000,) & 4MB & 35kb \\
page-only-8mb & paged-aggregated file with 4mb per pag8 &
\textasciitilde1\% & 1.5km & (10000,) & 8MB & 35kb \\
rechunked-4mb & page-aggregated and bigger chunk sizes &
\textasciitilde1\% & 10km & (100000,) & 4MB & 400kb \\
rechunked-8mb & page-aggregated and bigger chunk sizes &
\textasciitilde1\% & 10km & (100000,) & 8MB & 400kb \\
rechunked-8mb-kerchunk & kerchunk sidecar of the last paged-aggregated
file & N/A & 10km & (100000,) & 8MB & 400kb \\
\end{longtable}

@@

\includegraphics{figures/figure-4.png}

@@

\includegraphics{figures/figure-5.png}

@@

\subsection{Recommended access
patterns}\label{recommended-access-patterns}

In progress

\subsection{Recommended tooling
development}\label{recommended-tooling-development}

In progress

@@

\section{Discussion}\label{discussion}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Chunking shapes and sizes
\item
  Paged aggregation vs User block
\item
  Side effects on different access patterns, e.g.~Kerchunk
\end{enumerate}

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\usepackage{bookmark}

@@

\pandocbounded{\includegraphics[keepaspectratio]{figures/figure-1.png}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{figures/figure-2.png}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{figures/figure-3.png}}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1589}}
  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.3444}}
  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1457}}
  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.0993}}
  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.0728}}
  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.0728}}
  >{\raggedright\arraybackslash}p{(\linewidth - 12\tabcolsep) * \real{0.1060}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
prefix
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\% file size increase
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textasciitilde km per chunk
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
shape
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
page size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
avg\_chunk\_size
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
original & original file from ATL03 v6 (1gb and 7gb) & 0 & 1.5km &
(10000,) & N/A & 35kb \\
original-kerchunk & kerchunk sidecar of the original file & N/A & 1.5km
& (10000,) & N/A & 35kb \\
page-only-4mb & paged-aggregated file with 4mb per page &
\textasciitilde1\% & 1.5km & (10000,) & 4MB & 35kb \\
page-only-8mb & paged-aggregated file with 4mb per pag8 &
\textasciitilde1\% & 1.5km & (10000,) & 8MB & 35kb \\
rechunked-4mb & page-aggregated and bigger chunk sizes &
\textasciitilde1\% & 10km & (100000,) & 4MB & 400kb \\
rechunked-8mb & page-aggregated and bigger chunk sizes &
\textasciitilde1\% & 10km & (100000,) & 8MB & 400kb \\
rechunked-8mb-kerchunk & kerchunk sidecar of the last paged-aggregated
file & N/A & 10km & (100000,) & 8MB & 400kb \\
\end{longtable}

@@

\pandocbounded{\includegraphics[keepaspectratio]{figures/figure-4.png}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{figures/figure-5.png}}

@@

\subsection{Recommended Access
Patterns}\label{recommended-access-patterns}

As we saw in our benchmarks, efficient access to cloud optimized HDF5
files in cloud storage requires that we also optimize our access
patterns. The following recommendations focus on optimizing workflows
for Python users. However, these recommendations should be applicable
across programming languages. It's also worth mentioning that the HDF
Group aims to include some of these features in their roadmap.

\begin{itemize}
\tightlist
\item
  \textbf{Efficient Reads}: Efficiently reading cloud-hosted HDF5 files
  involves minimizing network requests and prioritizing large sequential
  reads. Configure chunk sizes between 1--10 MB to match the block sizes
  used in cloud object storage systems, ensuring meaningful data
  retrieval in each read. Avoid small chunks, as they cause excessive
  HTTP overhead and slower access speeds.
\item
  \textbf{Parallel Access}: Use parallel computing frameworks like
  \texttt{Dask} or multiprocessing to divide read operations across
  multiple processes or nodes. This alleviates the sequential access
  bottleneck caused by the HDF5 global lock, particularly in workflows
  accessing multiple datasets.
\item
  \textbf{Cache Management}: Implement caching for metadata to avoid
  repetitive fetches. Tools like \texttt{fsspec} or \texttt{h5coro}
  allow in-memory or on-disk caching for frequently accessed data,
  reducing latency during high-frequency
\item
  \textbf{Regional Access}: Operate workflows in the same cloud region
  as the data to minimize costs and latency. Cross-region data transfer
  is expensive and introduces significant delays. Where possible, deploy
  virtual machines close to the data storage region.
\end{itemize}

\subsection{Recommended Tooling
Development}\label{recommended-tooling-development}

To enable widespread and efficient use of HDF5 files in cloud
environments, it is crucial to develop robust tools across all major
programming languages. The HDF Group has expressed intentions to include
these features in their roadmap, ensuring seamless compatibility with
emerging cloud storage and computing standards. This section highlights
tooling strategies to support metadata indexing, driver enhancements,
and diagnostics, applicable to Python and other languages.

\begin{itemize}
\tightlist
\item
  \textbf{Enhanced HDF5 Drivers:} Improve drivers like \texttt{h5py} and
  \texttt{ROS3} to better handle cloud object storage's nuances, such as
  intelligent request batching and speculative reads. This mitigates
  inefficiencies caused by high-latency networks.
\item
  \textbf{Metadata Indexing:} Develop tools for pre-indexing metadata,
  similar to Kerchunk. These tools should enable clients to retrieve
  only necessary data offsets, avoiding full metadata reads and
  improving access times.
\item
  \textbf{Kerchunk-like Integration:} Extend Kerchunk to integrate
  seamlessly with analysis libraries like Xarray. This includes building
  robust sidecar files that efficiently map hierarchical datasets,
  enabling faster partial reads and enhancing cloud-native workflows.
\item
  \textbf{Diagnostic Tools:} Create tools for diagnostics and
  performance profiling tailored to cloud-optimized HDF5 files. These
  tools should identify bottlenecks in access patterns and recommend
  adjustments in configurations or chunking strategies.
\end{itemize}

@@

\subsection{Discussion and Further
Work}\label{discussion-and-further-work}

We believe that implementing cloud optimized HDF5 will greatly improve
downstream workflows that will unlock science in the cloud. We also
recognize that in order to get there, some key factors in the ecosystem
need to be addressed. Chunking strategies, adaptive caching and
automatic driver configurations should be developed to optimize
performance.

Efforts should expand multi-language support, creating universal
interfaces and libraries for broader adoption beyond Python.
Cloud-native enhancements must focus on optimizing HDF5 for distributed
systems and object storage, addressing egress costs, ease of use and
scalability. Finally, advancing ecosystem interoperability involves
setting integration standards and aligning with emerging trends such as
serverless and edge computing. These efforts, combined with community
collaboration, will modernize HDF5 to meet the challenges of evolving
data-intensive applications.

\subsubsection{Chunking Shapes and
Sizes}\label{chunking-shapes-and-sizes}

Optimizing chunk shapes and sizes is essential for efficient HDF5 usage,
especially in cloud environments:

\begin{itemize}
\tightlist
\item
  \textbf{Chunk Shape:} Align chunk dimensions with anticipated access
  patterns. For example, row-oriented queries benefit from row-aligned
  chunks.
\item
  \textbf{Chunk Size:} Use chunk sizes between 1--10 MB to match cloud
  storage block sizes. Larger chunks improve sequential access but
  require more memory. Smaller chunks support granular reads but may
  increase network overhead.
\end{itemize}

Finally, we recognize that this study has not been as extensive as it
could have been (cross language, multiple datasets) and yet we think we
ran into the key scenarios data producers will face when they start
producing cloud optimized HDf5 files. We think that there is room for
improvement and experimentation with various configurations based on
real-world scenarios is crucial to determine the best performance.
